Star-galaxy Classification Using Convolutional Neural Networks

### 2016-05-17 paper/main.tex
update draft paper

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[letterpaper,margin=0.65in]{geometry}
\usepackage[ansinew]{inputenc}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{color}
\usepackage{amssymb}
\usepackage[normalem]{ulem}
\usepackage{natbib}
\usepackage{url}
\usepackage{placeins}
\usepackage{times}

@@

\title[Deep Learning]{Deep Learning for Source Classification}

\author[E. J. Kim \& R. J. Brunner]{
  Edward J.~Kim$^1$\thanks{jkim575@illinois.edu},
  Robert J.~Brunner$^{2,3,4}$ \\
$^1$Department of Physics, University of Illinois, Urbana, IL 61801 USA\\
$^2$Department of Astronomy, University of Illinois, Urbana, IL 61801 USA\\
$^3$Department of Statistics, University of Illinois, Champaign, IL 61820 USA\\
$^4$National Center for Supercomputing Applications, Urbana, IL 61801 USA}

\begin{document}

\date{\today}

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}

\maketitle

\label{firstpage}
\begin{abstract}
Deep learning.
\end{abstract}

@@

\section{Methods}
  \label{section:methods}

@@

@@

>>>>

% Basic setup. Most papers should leave these options alone.
\documentclass[a4paper,fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
%\usepackage{newtxtext,newtxmath}
% Depending on your LaTeX fonts installation, you might get better results with one of these:
\usepackage{mathptmx}
%\usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{savesym}
\usepackage{amsmath}
\savesymbol{iint}
\usepackage{txfonts}
\restoresymbol{TXF}{iint}

\usepackage{graphicx}	% Including figure files
%\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[ansinew]{inputenc}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{color}
\usepackage[normalem]{ulem}
\usepackage{natbib}
\usepackage{url}
\usepackage{placeins}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

@@

\title[Star-galaxy Classification Using ConvNets]{Star-galaxy Classification Using Convolutional Neural Networks}

\author[E. J. Kim and R. J. Brunner]{
  Edward J.~Kim$^1$\thanks{jkim575@illinois.edu} and Robert J.~Brunner$^{2}$ \\
$^1$Department of Physics, University of Illinois, Urbana, IL 61801 USA\\
$^2$Department of Astronomy, University of Illinois, Urbana, IL 61801 USA}

% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

% Enter the current year, for the copyright statements etc.
\pubyear{2015}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
Most existing star-galaxy classifiers use the reduced summary information from catalogs,
requiring careful feature extraction and selection.
The latest advances in machine learning using deep convolutional neural networks
allow a machine to automatically learn the features directly from data,
minimizing need for input from human experts.
We present a star-galaxy classification framework that uses
a deep convolutional neural network (ConvNet) directly on the raw pixel values.
Using data from the Sloan Digital Sky Survey (SDSS) and
the Canada-France-Hawaii Telescope Lensing Survey (CFHTLenS),
we demonstrate that ConvNet is able to produce accurate and well-calibrated
probabilistic classifications that are competitive with
conventional machine learning techniques.
Future advances in deep learning may bring more success in currently ongoing and
forthcoming photometric surveys, such as the Dark Energy Survey (DES) and
the Large Synoptic Survey Telescope (LSST), because deep neural networks require
very little feature engineering by hand.
\end{abstract}

@@

Currently ongoing and forthcoming large-scale photometric surveys,
such as the Dark Energy Survey (DES) and the Large Synoptic Survey Telescope (LSST),
aim to collect photometric data for hundreds of millions to billions of
stars and galaxies.
Due to the sheer volume of data, it is not possible for human experts to
manually classify them,
and the separation of photometric catalogs into stars and galaxies has to be automated.
Furthermore, any classification approach must be probabilistic in nature.
A fully probabilistic classifier enables a user to adopt probability cuts
to obtain a pure sample for population studies,
or to optimize the allocation of observing time by selecting objects for follow-up.
However, ideally, the probability estimates themselves would be retained for all sources
and used in subsequent analyses to improve or enhance a particular
measurement~\citep{ross2011ameliorating,seo2012acoustic}.

With machine learning, we can use algorithms to automatically create
accurate source catalogs with well-calibrated posterior probabilities.
Machine learning techniques have been a popular tool
in many areas of astronomy~\citep{ball2007robust,ball2008robust,
banerji2010galaxy,
carrascokind2013tpz,carrascokind2014somz,carrascokind2014exhausting,
ivezic2014statistics,
kamdar2016machine1,kamdar2016machine2}.
Artificial neural networks were first applied to the problem of star-galaxy
classification in the work of \citet{odewahn1992automated},
which has become a core part of the astronomical image processing software
\texttt{SExtractor}~\citep{bertin1996sextractor}.
Other successfully implemented examples of applying machine learning to
the star-galaxy classification problem include decision 
trees~\citep{weir1995automated,suchkov2005census,ball2006robust,sevilla2015effect}
, Support Vector Machines~\citep{Fadely2012}, and
classifier combination strategies~\citep{kim2015hybrid}.

Almost all star-galaxy classifiers published in the literature use
the reduced summary information available from astronomical catalogs.
Constructing catalogs requires careful engineering and considerable domain expertise
to transform the raw pixel values of an image to suitable features,
such as magnitudes or shape information of an object.
In a branch of machine learning called deep learning,
features are not designed by human experts;
they are learned from data by deep neural networks.
Deep learning has been applied successfully to galaxy morphological
classification in Sloan Digital Sky Survey~\citep[SDSS;][]{dieleman2015rotation}
and Cosmic Assembly Near-infrared Deep Extragalactic Legacy
Survey~\citep[CANDELS;][]{huertas2015catalog} and to photometric redshift
estimation~\citep{hoyle2015measuring},
but it has never been applied to the problem of source classification.

In this paper, we present a star-galaxy classification framework
that uses a convolutional neural network (ConvNet) model directly on the images
from SDSS and Canada-France-Hawaii Telescope Lensing Survey (CFHTLenS).
We compare its performance with a standard machine learning technique
that uses the reduced summary information from catalogs,
and demonstrate that our ConvNet model is able to produce accurate and
well-calibrated probabilistic classifications with very little feature
engineering by hand.
In Section~\ref{section:data}, we describe the data sets used in this paper
and the pre-processing steps for preparing the image data sets.
We provide a brief overview of deep learning and ConvNets
in Section~\ref{section:deep_learning}, and
discuss our strategy for preventing overfitting in Section~\ref{section:overfitting}.
In Section~\ref{section:tpc}, we describe a state-of-the-art tree-based
machine learning algorithm, to which the performance of our ConvNet model is
compared.
We present the main results of our ConvNet model in
Section~\ref{section:results_and_discussion}, and
we outline our conclusions in Section~\ref{section:conclusions}.

@@

\subsection{Sloan Digital Sky Survey}
  \label{section:sdss}

The Sloan Digital Sky Survey~\citep[SDSS;][]{york2000sloan}
phases I--III obtained photometric data in five bands:
$u$, $g$, $r$, $i$, and $z$,
covering 14,555 square degrees, more than one-third of the entire sky.
The resulting catalog contains photometry of over 300 million stars galaxies
with a limiting magnitude of $r \approx 22$,
making the SDSS one of the largest sky surveys ever undertaken.
The SDSS also conducted an expansive spectroscopic follow-up of
more than three million stars and galaxies~\citep{eisenstein2011sdss}.
In this paper, we use a subset of the photometric and spectroscopic data
contained within the Data Release 12~\citep[DR12;][]{alam2015eleventh},
which is publicly available through the online CasJobs
server~\citep{li2008casjobs}~\footnote{http://skyserver.sdss.org/casjobs/}.

Using the CasJobs server, we randomly select 65,000 stars and galaxies.
We exclude some bad photometric observations as follows.
We consider only objects whose $r$ band magnitude is less than 40;
there are no warning flags in the spectroscopic measurement (\texttt{zWarning = 0});
the half-light radius in the $r$ band is less than 30 arc seconds
as measured by the exponential and de Vaucouleurs light profiles;
the error on the spectroscopic redshift measurement is less than 0.1; and
the spectroscopic redshift is less than 2.

To create training images,
we obtain the image FITS files for SDSS fields containing these objects
in five photometric bands: $u$, $g$, $r$, $i$, and $z$.
We use the astrometry information in the FITS headers
in the \texttt{Montage}~\footnote{http://montage.ipac.caltech.edu/} software
to align each image to the reference ($r$-band) image.
We then use \texttt{SExtractor} to find the pixel positions of
the 65,000 objects we have selected,
and to center each object in the cutout image.
We also convert the flux values to inverse hyperbolic sine magnitudes
\citep[also known as luptitudes;][]{lupton1999modified}.
Finally, in order to account for the effect of galactic dust,
galactic extinction corrections in magnitudes are applied
following \cite{schlegel1998maps}.
In the end, we have cutout images of size $48\times48$ pixels
with luptitude values in each pixel.
We note that we have experimented with increasing the pixel dimensions to
$60\times60$ and $72\times72$ pixels, but do not find noticeable
improvement in the performance of our model.

In the end, we have 17 344 stars and 47 656 galaxies available
for the training and testing processes.
We randomly split the objects into training, validation, and test sets
of 40 000, 10 000, and 15 000, respectively.
We note that we perform a blind test, and the test set is not used in any
to train or calibrate the algorithms.

\subsection{Canada-France-Hawaii Telescope Lensing Survey}

We also use photometric data from
the Canada-France-Hawaii Telescope Lensing Survey
\cite[CFHTLenS\footnote{http://www.cfhtlens.org/};][]
{heymans2012cfhtlens,erben2013cfhtlens,hildebrandt2012cfhtlens}.
This catalog consists of more than twenty five million objects
with a limiting magnitude of $i_{\text{AB}} \approx 25.5$. 
It covers a total of 154 square degrees
in the four fields (named W1, W2, W3, and W4)
of CFHT Legacy Survey~\citep[CFHTLS;][]{gwyn2012canada}
observed in the five photometric bands:
$u$, $g$, $r$, $i$, and $z$.

We have cross-matched reliable spectroscopic galaxies from
the Deep Extragalactic Evolutionary Probe Phase 2~
\citep[DEEP2;][]{davis2003science,newman2013deep2},
the Sloan Digital Sky Survey Data Release 10~\citep[SDSS-DR10]{alam2015eleventh},
the VIsible imaging Multi-Object Spectrograph (VIMOS)
Very Large Telescope (VLT) Deep Survey~
\citep[VVDS;][]{le2005vimos,garilli2008vimos}, and
the VIMOS Public Extragalactic Redshift
Survey~\citep[VIPERS;][]{garilli2014vimos}.
We have selected only sources with very secure
redshifts and no bad flags (quality flags -1, 3, and 4 for DEEP2;
quality flag 0 for SDSS; quality flags 3, 4, 23, and 24 for VIPERS
and VVDS).

We obtain FITS images for each 1 square degree CFHTLenS pointing
that contains objects with spectroscopic labels.
We create cutout images of size $96\times96$ pixels by using a similar method
to that described in Section~\ref{section:sdss}.
Finally, images are downscaled to $48\times48$ pixels
to reduce reduce computational cost.

In the end, we have 8 545 stars and 57 843 galaxies available
for the training and testing processes.
We randomly split the objects into training, validation, and test sets of size
40 000, 10 000, and 13 278, respectively.

\section{Deep Learning}
  \label{section:deep_learning}

\subsection{Neural Networks}

\begin{figure*}
  \centering
  \begin{subfigure}[]{0.49\linewidth}
    \centering
    \includegraphics[width=0.6\textwidth]{figures/neuron.pdf}
    \caption{}
    \label{fig:neuron_a}
  \end{subfigure}
  \begin{subfigure}[]{0.49\linewidth}
    \centering
    \includegraphics[width=0.6\textwidth]{figures/network.pdf}
    \caption{}
    \label{fig:neuron_b}
  \end{subfigure}
  \caption{
    (a) A mathematical model of a biological neuron.
    (b) A schematic diagram of a neural network with one hidden layer
    of two neurons and one output layer with two neurons.
    }
\end{figure*}

A neuron in the human brain receives signals from other neurons through
synaptic connections.
If the combination of these signals exceeds a certain threshold,
the neuron will fire and send a signal to other neurons.
Intelligence is believed to be the collective effect of
approximately $10^{11}$ neurons firing.
An artificial neuron in most artificial neural networks is represented
as a mathematical function that models a biological neural structure
(Figure~\ref{fig:neuron_a}).
Let $\mathbf{x}$ be a vector of inputs to a given neuron,
$\mathbf{w}$ be a vector of weights, and
$\mathbf{b}$ be the bias.
Then, the output of the neuron is
\begin{equation}
  y = \sigma \left( \mathbf{w} \cdot \mathbf{x} + b \right),
  \label{eq:neuron_output}
\end{equation}
where $\sigma$ is the activation function (or \textit{non-linearity}).

The sigmoid function, $\sigma(x)=1/\left(1+e^{-x}\right)$,
has frequently been used as activation function, but it has recently fallen out of favor,
and the hyperbolic tangent function, $\sigma(x)=\tanh(x)$ is almost always used instead.
The most popular non-linearity at present, however, is the rectified linear unit
\citep[ReLU;][]{nair2010rectified}, $\sigma(x)=\max(0, x)$,
ReLUs generally allow much faster
training of deep neural networks with many layers.
However, ReLU units can sometimes result in dead neurons whose output is always zero.
To mitigate this problem, we use leaky ReLUs that have a small negative slope
in the negative region,
\begin{equation}
  \sigma(x) =
    \begin{cases}
      x & \mbox{if } x \geq 0 \\
      0.01x & \mbox{if } x < 0.
    \end{cases}
\end{equation}

Many deep learning models use feedforward neural network architectures with multiple layers,
where each neuron in one layer is connected to the neurons of the subsequent layer.
A schematic representation is shown in Figure~\ref{fig:neuron_b}.
All layers except the input and output layers are conveniently called hidden layers.
With multiple hidden layers, deep neural networks can model complex nonlinear
relationships, and they have had great success in multimedia tasks where humans traditionally
outperformed computers: speech recognition, language processing, and image classification.

The goal of learning is to find a set of weights and biases such that
the output from the network $\mathbf{y}=\left\{ y_1, y_2, \dots, y_n \right\}$
approximates the desired output
$\hat{\mathbf{y}}=\left\{\hat{y}_1, \hat{y}_2, \dots, \hat{y}_n \right\}$
as closely as possible for all input $\mathbf{x}$.
We can formulate this as the minimization of a loss function $L(\mathbf{x},\hat{\mathbf{y}})$
over the training data.
In this work, we use \textit{cross-entropy},
\begin{equation}
  L(y, \hat{y}) = -\hat{y} \log_2 y - (1 - \hat{y}) \log_2(1 - y),
\end{equation}
where $\hat{y}$ is the actual truth value (\eg 0 or 1)
and $y$ is the probability prediction made by the model.
We compute the loss function by taking the average of all cross-entropies in the sample.
Given $N$ samples, the loss function becomes
\begin{equation}
  L(y, \hat{y}) = - \frac{1}{N} \sum_{i=1}^{N} \hat{y}_i  \log_2 y_i
    + (1 - \hat{y}_i)  \log_2 (1 - \hat y_i).
\end{equation}
where $\hat{y}_i$ is the truth value and probability estimate of the $i$-th data,
respectively.

To find the weights $\mathbf{w}$ and biases $\mathbf{b}$ which minimize the loss,
we use a technique called \textit{gradient descent},
where we the following rules to update the parameters in each layer $j$:
\begin{align}
  \mathbf{w}_j &\rightarrow
  \mathbf{w}^{\prime}
  = \mathbf{w}_j - \eta \frac{\partial L}{\partial \mathbf{w}_j} \\
  \mathbf{b}_j &\rightarrow
  \mathbf{b}^{\prime}
  = \mathbf{b}_j - \eta \frac{\partial L}{\partial \mathbf{b}_j},
  \label{eq:gradeitn_descent}
\end{align}
where $\eta$ is a small, positive number known as the \textit{learning rate}.
A common approach to speed up training is to split the training data
into mini-batches and compute the gradients in
Equation~\ref{eq:gradeitn_descent} for each mini-batch separately instead of
the entire training data.
This results in much faster convergence, as training examples are usually correlated
and the gradient computed from a randomly chosen mini-batch is a good approximation
of the overall gradient.
We define an \textit{epoch} as a single, complete pass through the training data,
and full training usually requires many epochs.
At the end of each epoch, we evaluate the loss function on the validation set,
and the model that minimizes the loss is chosen as the best model.

\subsection{Convolutional Neural Networks}

The Convolutional neural network~\citep[ConvNet;][]{fukushima1980neocognitron,lecun1998gradient}
is a type of deep, feedforward neural network
that has recently become the dominant approach in the computer vision community.
In a typical ConvNet, the first few stages are composed of two types of layers: 
convolutional layers and pooling layers.

The input to a convolutional layer is an image with a certain number of color channels
or \textit{feature maps}.
To produce output feature maps, we convolve each feature map with a set of weights
called \textit{filters},
and apply a non-linearity such as ReLU to the weighted sum of these convolutions.
Different feature maps use different sets of filters,
but all neurons in a feature map share the same set of filters.
Mathematically, we replace the dot product in Equation~\ref{eq:neuron_output}
with a sum of convolutions. The $k$-th feature map is given by
\begin{equation}
\mathbf{y}^k = \sigma \left( \sum_{j} \mathbf{w}_j^k \ast \mathbf{x}_j + b^k \right),
\end{equation}
where we sum over the set of filters,
$\ast$ is the convolution operator, and $\mathbf{w}_j^k$ represent the filters.

A typical pooling layer computes the maximum of a local $2\times2$ patch in a feature map.
Since the pooling layer aggregates the activations of neighboring units from the previous layer,
it reduces the dimensionality of the feature maps and
makes the model invariant to small shifts and distortions.
Two or more layers of convolution and pooling are stacked,
followed by more convolutional and fully-connected layers.

\subsection{Neural Network Architecture}

\begin{table*}
  \centering
  \caption{Summary of hyperparameters. Note that pooling layers have no learnable parameters.}
  \label{table:hyperparamters}
  \begin{tabular}{ccccccc}
    \hline
    type            & filters & filter size & padding & non-linearity & initial weights & initial biases \\
    \hline
    convolutional   & 32          & $5\times5$  & -       & leaky ReLU    & orthogonal      & 0.1            \\
    convolutional   & 32          & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    pooling         & -           & $2\times2$  & -       & -             & -               & -              \\
    convolutional   & 64          & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    convolutional   & 64          & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    convolutional   & 64          & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    pooling         & -           & $2\times2$  & -       & -             & -               & -              \\
    convolutional   & 128         & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    convolutional   & 128         & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    convolutional   & 128         & $3\times3$  & 1       & leaky ReLU    & orthogonal      & 0.1            \\
    pooling         & -           & $2\times2$  & -       & -             & -               & -              \\
    fully-connected & 2048        & -           & -       & leaky ReLU    & orthogonal      & 0.01            \\
    fully-connected & 2048        & -           & -       & leaky ReLU    & orthogonal      & 0.01            \\
    fully-connected & 2           & -           & -       & softmax       & orthogonal      & 0.01            \\
    \hline
  \end{tabular}
\end{table*}

We present the overall architecture of our ConvNet model in Table~\ref{table:hyperparamters}.
The network consists of eleven trainable layers.
The first convolutional layer filters the $5\times44\times44$ input image
(\ie $44\times44$ images in five bands $ugriz$) with 32 square filters of size $5\times5\times5$.
The leaky ReLU non-linearity is applied to the output of the first convolutional layer
(and all subsequent layers), and the second convolutional layer filters it
with 32 filters of $32\times3\times3$.
In the second convolutional layer (and all subsequent convolutional layers),
we pad the input with zeros spatially on the border
(\ie the zero-padding is 1 pixel for $3\times3$ convolutional layers)
such that the spatial resolution is preserved after convolution.
Max-pooling with filters of size $2\times2$ follows the second convolutional layer.
A stack of six additional convolutional layers, all with filters of size $3\times3$,
is followed by three fully-connected layers.
The first two fully-connected layers have 2048 channels each,
and the third performs binary classification.

The output of the final fully-connected layer is fed to a softmax function.
The softmax function is given by
\begin{equation}
  P (y_i \mid \mathbf{x}) = 
    \frac{ e^{ \mathbf{x} \cdot \mathbf{w}_{y_i} } }
    { \sum_{k} e^{ \mathbf{x} \cdot \mathbf{w}_k } },
\end{equation}
where we interpret its output as the posterior probability that an object is a galaxy
(or a star).
We note that we could also try to solve a regression problem, \eg
by normalizing the output values that the network produces for each class.
However, we find that solving a regression problem instead of using
the softmax function results in significantly worse performance.

In addition to hyperparamter optimization, we have explored more than 100 different architectures
and have found that the architecture described in this section provides
the best performance on the validation data.
For example, the architecture of \cite{krizhevsky2012imagenet}
uses relatively large receptive fields ($11\times11$) in the first convolutional layers.
\citet{zeiler2014visualizing} and \cite{dieleman2015rotation}
also use large receptive fields of $7\times7$ and $6\times6$
in the first convolution layer, respectively.
However, we find that using a receptive field larger than $5\times5$ in the first
convolutional layer leads to worse performance.
This result is in agreement with \citet{simonyan2014very},
which features an extremely homogeneous architecture that only performs
$3\times3$ convolutions.
Using a large receptive field instead of a stack of multiple $3\times3$ convolutions
leads to a shallower network, and it is often preferable to increase
the depth by using smaller receptive fields.
However, we find that replacing the first layer with
a stack of two $3\times3$ convolutional layers increases the validation error,
but we still follow \citet{simonyan2014very} and add many
$3\times3$ convolutions (with zero-padding) in the remaining layers.

\subsection{Training}

The choice of momentum, learning rate, and initial weights is crucial
for achieving high predictive performance and speeding up the learning
process~\citep{sutskever2013importance}.
To train our models, we use mini-batch gradient descent with a batch size of 128
and Nesterov momentum~\citep{bengio2013advances} of 0.9.
We initialize the learning rate $\eta$ at 0.003 for all layers and
decrease it linearly with the number of epochs 
from 0.003 to 0.0001 over 750 epochs.
We also initialize the weights in each layer with random orthogonal initial 
conditions~\citep{saxe2013exact}.
We use a slightly positive value $b=0.1$ for all biases.
We find initializing biases to a small constant value helps eliminate dead neurons
by ensuring that all ReLU neurons fire in the beginning.
We note that our network takes about 40 hours to train the networks on a NVIDIA Tesla K40 GPU.

\section{Reducing Overfitting}
  \label{section:overfitting}

Our convolution neural network has $11\times10^6$ learnable parameters,
while there are only $5\times10^4$ images in the training set.
As a result, the model is very likely to \emph{overfit} without regularization.
In this section, we describe the techniques used to combat overfitting.

\subsection{Data Augmentation}
  \label{sec:data_augmentation}

One common method to combat overfitting is to artificially increase
the number of training data by using label-preserving transformations.
Each image is transformed as follows:
\begin{itemize}
\item{Rotation: 
Rotating an image does not change whether the object is a star or a galaxy.
We exploit this rotational symmetry and randomly rotate each image by a multiple of
$90^{\circ}$. }

\item{Reflection:
We flip each image horizontally with a probability of 0.5 to exploit mirror symmetry. }
\item{Translation:
We also have translational symmetry in the images.
Given an image of size $48\times48$ pixels, we extract a random contiguous crop
of size $44\times44$.
Each cropping is equivalent to randomly shifting a $44\times44$ image by up to 4 pixels
vertically and horizontally. }
\item{Gaussian noise:
We introduce random Gaussian noise to each pixel values
by using a similar method to \cite{krizhevsky2012imagenet}. }
\end{itemize}
These data augmentation schemes add almost no computational cost,
as they are performed on the CPU while the GPU is training the ConvNets on images.

\subsection{Dropout}

We use a regularization technique called dropout~\citep{hinton2012improving}
in the fully-connected layers.
Dropout consists of randomly setting to zero
the output of each hidden neuron of the previous layer with probability 0.5.
The weights of the remaining neurons are multiplied by 0.5 to preserve
the scale of input values to the next layer.
Since a neuron can be removed at any time, it cannot rely on the presence of other neurons
in the same layer and is forced to learn more robust features.

\subsection{Model Combination}
  \label{sec:bmc}

To make final classifications,
we use our ConvNet model to make 64 sets of predictions for
64 transformations of the input images:
4 rotations, 4 horizontal translations, and 4 vertical translations
(with random horizontal reflections).
Although we use an identical network architecture for all transformations,
we consider each set of predictions as separate results from different models.
Finally, we use a model combination technique known as
Bayesian Model Combination~\citep[BMC;][]{Monteith2011}, which
uses Bayesian principles to generate an ensemble combination of different models.
Although the data augmentation step~\ref{sec:data_augmentation} should
make our ConvNet model invariant to these types of transformations,
we find that applying BMC still results in a significant decrease in validation error.

The posterior probability that a source is a star is given by
\begin{equation} \label{eq:p_star_bmc}
  P \left(S | \bmath{x}, \bmath{D}, \bmath{M}, \bmath{E} \right)
  = \sum_{e \in \bmath{E}} P \left(S | \bmath{x}, \bmath{M}, e \right)
  P \left(e | \bmath{D} \right),
\end{equation}
where $\bmath{D}$ is the data set,
and $e$ is an element in the ensemble space $\bmath{E}$ of possible model combinations.
By Bayes' Theorem, the posterior probability of $e$ given $\bmath{D}$ is given by
\begin{equation} \label{eq:p_ensemble}
  P \left(e | \bmath{D} \right)
  = \frac{P \left(e \right)}{P \left(\bmath{D} \right)}
  \prod_{d \in \bmath{D}} P \left( d | e \right)
  \propto P \left(e\right) \prod_{d \in \bmath{D}} P \left(d | e \right).
\end{equation}
Here, $P\left(e \right)$ is the prior probability of $e$,
which we assume to be uniform.
The product of $P\left(d | e\right)$ is
over all individual data $d$ in the training data $\bmath{D}$,
and $P\left(\bmath{D}\right)$ is merely a normalization factor
and not important.
For probabilistic classifiers,
we can directly use the probabilistic predictions
and write Equation~\ref{eq:p_ensemble} as
\begin{equation} \label{eq:p_ensemble_prob}
  P \left( e | \bmath{D} \right) \propto 
  P \left( e \right) \prod_{i=0}^{N-1}
  \hat{y}_i y_i + 
  \left(1 - \hat{y}_i\right) \left(1 - y_i\right).
\end{equation}

Although the space $\bmath{E}$ of potential model combinations
is in principle infinite,
we can produce a reasonable finite set
of potential model combinations by using sampling techniques.
In our implementation,
the weights of each combination of the base classifiers
is obtained by sampling from a Dirichlet distribution.
We first set all alpha values of a Dirichlet distribution to unity.
We then sample this distribution $q$ times
to obtain $q$ sets of weights.
For each combination,
we assume a uniform prior and
calculate $P\left(e|\bmath{D}\right)$ using
Equation~\ref{eq:p_ensemble_prob}.
We select the combination with the highest $P\left(e|\bmath{D}\right)$,
and update the alpha values by
adding the weights of the most probable combination
to the current alpha values.
The next $q$ sets of weights are drawn
using the updated alpha values.

We continue the sampling process until
we reach a predefined number of combinations,
and finally use Equation~\ref{eq:p_star_bmc} to compute
the posterior probability that a source is a star (or a galaxy).
In this paper, we use a $q$ value of three,
and 1,000 model combinations are considered.

\section{Trees for Probabilistic Classifications}
  \label{section:tpc}

To compare the performance of ConvNets with
machine learning algorithms that use standard photometric features,
we use a machine learning framework called
Trees for Probabilistic Classifications (TPC).
Although a complete description of TPC is beyond the scope of this paper,
we provide a brief description in this section.
For more details, we refer the reader to
\cite{carrascokind2013tpz} and \cite{kim2015hybrid}.

TPC is a parallel, supervised machine learning algorithm
that uses prediction trees and random forest 
techniques~\citep{breiman1984classification, breiman2001random}
to produce a star-galaxy classification.
TPC is a part of a publicly available software package called
\textsc{MLZ}\footnote{http://lcdm.astro.illinois.edu/code/mlz.html}
(Machine Learning for Photo-$z$).

TPC uses classification trees,
a type of prediction trees that are designed to
provide a classification or predict a discrete category.
Prediction trees are built by asking a sequence of questions
that recursively split the data into branches
until a terminal leaf is created
that meets a stopping criterion
(\eg a minimum leaf size).
The optimal split dimension is decided by
choosing the attribute that maximizes
the \textit{Information Gain} ($I_G$), which is defined as

\begin{equation} \label{eq:information_gain}
  I_G \left(D_{\rmn{node}}, X\right)
  = I_d \left( D_{\rmn{node}} \right)
  - \sum_{x \in \rmn{values}(X)}
  \frac{|D_{\rmn{node}, x}|}{|D_{\rmn{node}}|}
  I_d \left( D_{\rmn{node}, x} \right),
\end{equation}

\noindent
where $D_{\rmn{node}}$ is the training data in a given node,
$X$ is one of the possible dimensions (\eg magnitudes or colors)
along which the node is split, and
$x$ are the possible values of a specific dimension $X$.
$|D_{\rmn{node}}|$ and $|D_{\rmn{node}, x}|$ are the size of the total training data
and the number of objects in a given subset $x$ within the current node,
respectively.
$I_d$ is the impurity degree index,
and TPC can calculate $I_d$
from any of the three standard different impurity indices:
\textit{information entropy}, \textit{Gini impurity},
and \textit{classification error}.
In this work, we use the information entropy,
which is defined similarly to the thermodynamic entropy:

\begin{equation}
  I_d \left( D \right)
  = - f_g \log_{2} f_g - \left(1 - f_g\right) \log_{2} \left(1 - f_g\right),
\end{equation}

\noindent
where $f_g$ is the fraction of galaxies in the training data.
At each node in our tree,
we scan all dimensions to identify the split point that
maximizes the information gain as defined by Equation~\ref{eq:information_gain},
and select the attribute that maximizes the impurity index overall.

In a technique called random forest,
we create bootstrap samples
(\ie $N$ randomly selected objects with replacement)
from the input training data
by sampling repeatedly from the magnitudes and colors
using their measurement errors.
We use these bootstrap samples to construct
multiple, uncorrelated prediction trees
whose individual predictions are aggregated to produce
a star-galaxy classification for each source.

We train two TPC models on the SDSS data set using different sets of
attributes.
The first model, which we denote $\rmn{TPC}_{\rmn{phot}}$, is trained with
only nine photometric attributes:
the extinction-corrected model magnitudes in five bands
($u$, $g$, $r$, $i$, $z$) and their corresponding colors
($u-g$, $g-r$, $r-i$, $i-z$).
The second model, which we denote $\rmn{TPC}_{\rmn{morph}}$, is trained with
the concentration parameter in each band in addition to the magnitudes and
colors, for a total of fifteen dimensions.
The concentration is defined as the difference between the cmodel
and PSF magnitude,
\ie $\rmn{concentration}\equiv\rmn{psfMag}-\rmn{cModelMag}$.
The SDSS pipeline uses a parametric method based on the concentration,
\ie an object is classified as a galaxy if $\rmn{concentration}>0.145$.
We find that the concentration is an excellent morphological feature
for star-galaxy separation, and including more morphological features
does not show noticeable improvement in performance.
The concentration is a good example of carefully handcrafted feature extraction,
and we show in Section~\ref{section:results_and_discussion}
that ConvNets do not require such feature engineering.

We also train two models on the CFHTLenS data set.
$\rmn{TPC}_{\rmn{phot}}$ is trained with 
the five magnitudes and their corresponding colors,
$u$, $g$, $r$, $i$, $z$, $u-g$, $g-r$, $r-i$, and $i-z$.
Since the CFHTLenS catalog does not provide the concentration parameter,
$\rmn{TPC}_{\rmn{morph}}$ uses
\texttt{SExtractor}'s \texttt{FLUX\_RADIUS} (the half-light radius),
\texttt{A\_WORLD} (the semi-major axis), \texttt{B\_WORLD} (the semi-minor axis)
for morphological features, in addition to
the five magnitudes and their corresponding colors,
for a total of twelve dimensions.

@@

In this section, we present the classification performance of
our ConvNet model and compare it with the performance of TPC.

\subsection{Classification Metrics}

Probabilistic classification models can be considered as
functions that output a probability estimate of each source
to be in one of the classes (\eg a star or a galaxy).
The simplest way to way to use the probability output is to
convert it into a class label by using a threshold (a probability cut).
Although many studies often choose the threshold by setting it to a fixed value,
\eg $p_\rmn{cut} = 0.5$ \citep[\eg][]{henrion2011bayesian, Fadely2012},
choosing $0.5$ as a threshold is not the best choice
for an unbalanced data set, where galaxies outnumber stars.
Furthermore, setting a fixed threshold ignores the operating condition 
(\eg science requirements, stellar distribution, misclassification costs)
where the model will be applied.

\subsubsection{Receiver Operating Characteristic Curve}

When we have no information about the operating condition
when evaluating the performance of classifiers,
there are effective tools such as
the Receiver Operating Characteristic (ROC) curve
\citep*{swets2000better}.
An ROC curve is a graphical plot that illustrates the true positive rate
versus the false positive rate of a binary classifier
as its classification threshold is varied.
The Area Under the Curve (AUC) summarizes the curve information
in a single number,
and can be used as an assessment of the overall performance.

\subsubsection{Completeness and Purity}

In astronomical applications,
the operating condition usually translates to
the completeness and purity requirements of the star or galaxy sample.
We define the galaxy \textit{completeness}
$c_g$ (also known as recall or sensitivity) as
the fraction of the number of true galaxies classified as galaxies
out of the total number of true galaxies,
\begin{equation}
c_g = \frac{N_g}{N_g + M_g},
\end{equation}

\noindent
where $N_g$ is the number of true galaxies classified as galaxies,
and $M_g$ is the number of true galaxies classified as stars.
We define the galaxy \textit{purity} $p_g$ (also known as precision
or positive predictive value)
as the fraction of the number of true galaxies classified as galaxies
out of the total number of objects classified as galaxies, 
\begin{equation}
p_g = \frac{N_g}{N_g + M_s},
\end{equation}

\noindent
where $M_s$ is the number of true stars classified as galaxies.
Star completeness and purity are defined in a similar manner.
